\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\ Beyond repeated measures ANOVA and Linear Mixed Models\\ \strut \\ SUPPLEMENTARY MATERIALS: APPENDIX

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/SIM_Appendix.tex
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

%\lstset{
%  breaklines=true,
%  stringstyle=\ttfamily,
%  backgroundcolor=\color{gray}
%}
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
Beyond repeated measures ANOVA and Linear Mixed Models\\
~\\
SUPPLEMENTARY MATERIALS: APPENDIX}
\date{\vspace{-2.5em}}

@@

For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment. To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:

\begin{lstlisting}[language=R]
m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
\end{lstlisting}

This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots. The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}. The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}. A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.

@@

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists. The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing. However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends. Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B. If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D). Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.

@@

On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3. Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-roblyer2011}{}%
\CSLLeftMargin{{[}1{]} }
\CSLRightInline{D. Roblyer, S. Ueda, A. Cerussi, W. Tanamai, A. Durkin, R. Mehta, D. Hsiang, J.A. Butler, C. McLaren, W.-P. Chen, B. Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences. 108 (2011) 14626--14631. \url{https://doi.org/10.1073/pnas.1013103108}.}

\leavevmode\hypertarget{ref-tank2020}{}%
\CSLLeftMargin{{[}2{]} }
\CSLRightInline{A. Tank, H.M. Peterson, V. Pera, S. Tabassum, A. Leproux, T. O'Sullivan, E. Jones, H. Cabral, N. Ko, R.S. Mehta, B.J. Tromberg, D. Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research. 22 (2020). \url{https://doi.org/10.1186/s13058-020-01262-1}.}

\leavevmode\hypertarget{ref-pavlov2018}{}%
\CSLLeftMargin{{[}3{]} }
\CSLRightInline{M.V. Pavlov, T.I. Kalganova, Y.S. Lyubimtseva, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics. 23 (2018) 1. \url{https://doi.org/10.1117/1.jbo.23.9.091410}.}

\leavevmode\hypertarget{ref-demidov2018}{}%
\CSLLeftMargin{{[}4{]} }
\CSLRightInline{V. Demidov, A. Maeda, M. Sugita, V. Madge, S. Sadanand, C. Flueraru, I.A. Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports. 8 (2018). \url{https://doi.org/10.1038/s41598-017-18635-w}.}

\leavevmode\hypertarget{ref-ritter2001}{}%
\CSLLeftMargin{{[}5{]} }
\CSLRightInline{G. Ritter, L. Cohen, C. Williams, E. Richards, L. Old, S. Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}. {61} (2001) 6851--6859.}

\leavevmode\hypertarget{ref-roth2017}{}%
\CSLLeftMargin{{[}6{]} }
\CSLRightInline{E.M. Roth, A.C. Goldberg, A.L. Catapano, A. Torri, G.D. Yancopoulos, N. Stahl, A. Brunet, G. Lecorps, H.M. Colhoun, Antidrug antibodies in patients treated with alirocumab, New England Journal of Medicine. 376 (2017) 1589--1590. \url{https://doi.org/10.1056/nejmc1616623}.}

\leavevmode\hypertarget{ref-jones2018}{}%
\CSLLeftMargin{{[}7{]} }
\CSLRightInline{J.D. Jones, H.E. Ramser, A.E. Woessner, K.P. Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology. 1 (2018). \url{https://doi.org/10.1038/s42003-018-0206-4}.}

\leavevmode\hypertarget{ref-skala2010}{}%
\CSLLeftMargin{{[}8{]} }
\CSLRightInline{M.C. Skala, A. Fontanella, L. Lan, J.A. Izatt, M.W. Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics. 15 (2010) 011112. \url{https://doi.org/10.1117/1.3285584}.}

\leavevmode\hypertarget{ref-greening2018}{}%
\CSLLeftMargin{{[}9{]} }
\CSLRightInline{G.J. Greening, K.P. Miller, C.R. Spainhour, M.D. Cato, T.J. Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express. 9 (2018) 2871. \url{https://doi.org/10.1364/boe.9.002871}.}

\leavevmode\hypertarget{ref-sio2016}{}%
\CSLLeftMargin{{[}10{]} }
\CSLRightInline{T.T. Sio, P.J. Atherton, B.J. Birckhead, D.J. Schwartz, J.A. Sloan, D.K. Seisler, J.A. Martenson, C.L. Loprinzi, P.C. Griffin, R.F. Morton, J.C. Anders, T.J. Stoffel, R.E. Haselow, R.B. Mowat, M.A.N. Wittich, J.D. Bearden, R.C. Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer. 24 (2016) 3847--3855. \url{https://doi.org/10.1007/s00520-016-3213-3}.}

\leavevmode\hypertarget{ref-kamstra2015}{}%
\CSLLeftMargin{{[}11{]} }
\CSLRightInline{J.I. Kamstra, P.U. Dijkstra, M. van Leeuwen, J.L.N. Roodenburg, J.A. Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology. 51 (2015) 548--555. \url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.}

\leavevmode\hypertarget{ref-wagenmakers2008}{}%
\CSLLeftMargin{{[}12{]} }
\CSLRightInline{E.-J. Wagenmakers, M. Lee, T. Lodewyckx, G.J. Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer New York, 2008: pp. 181--207. \url{https://doi.org/10.1007/978-0-387-09612-4_9}.}

\leavevmode\hypertarget{ref-gueorguieva2004}{}%
\CSLLeftMargin{{[}13{]} }
\CSLRightInline{R. Gueorguieva, J.H. Krystal, Move over {ANOVA}, Archives of General Psychiatry. 61 (2004) 310. \url{https://doi.org/10.1001/archpsyc.61.3.310}.}

\leavevmode\hypertarget{ref-schober2018}{}%
\CSLLeftMargin{{[}14{]} }
\CSLRightInline{P. Schober, T.R. Vetter, Repeated measures designs and analysis of longitudinal data, Anesthesia {\&} Analgesia. 127 (2018) 569--575. \url{https://doi.org/10.1213/ane.0000000000003511}.}

\leavevmode\hypertarget{ref-pinheiro2006}{}%
\CSLLeftMargin{{[}15{]} }
\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science; Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}

\leavevmode\hypertarget{ref-vishwanath2009}{}%
\CSLLeftMargin{{[}16{]} }
\CSLRightInline{K. Vishwanath, H. Yuan, W.T. Barry, M.W. Dewhirst, N. Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia. 11 (2009) 889--900. \url{https://doi.org/10.1593/neo.09580}.}

\leavevmode\hypertarget{ref-dennis2019}{}%
\CSLLeftMargin{{[}17{]} }
\CSLRightInline{B. Dennis, J.M. Ponciano, M.L. Taper, S.R. Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}, Frontiers in Ecology and Evolution. 7 (2019). \url{https://doi.org/10.3389/fevo.2019.00372}.}

\leavevmode\hypertarget{ref-wang2019}{}%
\CSLLeftMargin{{[}18{]} }
\CSLRightInline{B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (2019) e100081. \url{https://doi.org/10.1136/gpsych-2019-100081}.}

\leavevmode\hypertarget{ref-liu2010}{}%
\CSLLeftMargin{{[}19{]} }
\CSLRightInline{C. Liu, T.P. Cripe, M.-O. Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy. 18 (2010) 1724--1730. \url{https://doi.org/10.1038/mt.2010.127}.}

\leavevmode\hypertarget{ref-halsey2015}{}%
\CSLLeftMargin{{[}20{]} }
\CSLRightInline{L.G. Halsey, D. Curran-Everett, S.L. Vowler, G.B. Drummond, The fickle p value generates irreproducible results, Nature Methods. 12 (2015) 179--185. \url{https://doi.org/10.1038/nmeth.3288}.}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
Beyond repeated measures ANOVA and Linear Mixed Models\\
\strut \\
SUPPLEMENTARY MATERIALS: APPENDIX}
\date{\vspace{-2.5em}}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

%\lstset{
%  breaklines=true,
%  stringstyle=\ttfamily,
%  backgroundcolor=\color{gray}
%}
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment. To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means. Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:

\begin{lstlisting}[language=R]
gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
\end{lstlisting}

This syntax specifies that \passthrough{\lstinline!gam\_02!} (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots. The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}. The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}. A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.

@@

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists. The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing. However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends. Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B. If 40\% of the observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D). Although the confidence intervals have increased for both smooths, the model still shows different trends with as few as 4 observations per group at certain time points.

@@

On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths shows the same trend of the full dataset. Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that the Control Group has higher \(\mbox{StO}_2\) before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-roblyer2011}{}}%
\CSLLeftMargin{{[}1{]} }
\CSLRightInline{D. Roblyer, S. Ueda, A. Cerussi, W. Tanamai, A. Durkin, R. Mehta, D. Hsiang, J.A. Butler, C. McLaren, W.-P. Chen, B. Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences. 108 (2011) 14626--14631. \url{https://doi.org/10.1073/pnas.1013103108}.}

\leavevmode\vadjust pre{\hypertarget{ref-tank2020}{}}%
\CSLLeftMargin{{[}2{]} }
\CSLRightInline{A. Tank, H.M. Peterson, V. Pera, S. Tabassum, A. Leproux, T. O'Sullivan, E. Jones, H. Cabral, N. Ko, R.S. Mehta, B.J. Tromberg, D. Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research. 22 (2020). \url{https://doi.org/10.1186/s13058-020-01262-1}.}

\leavevmode\vadjust pre{\hypertarget{ref-pavlov2018}{}}%
\CSLLeftMargin{{[}3{]} }
\CSLRightInline{M.V. Pavlov, T.I. Kalganova, Y.S. Lyubimtseva, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics. 23 (2018) 1. \url{https://doi.org/10.1117/1.jbo.23.9.091410}.}

\leavevmode\vadjust pre{\hypertarget{ref-demidov2018}{}}%
\CSLLeftMargin{{[}4{]} }
\CSLRightInline{V. Demidov, A. Maeda, M. Sugita, V. Madge, S. Sadanand, C. Flueraru, I.A. Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports. 8 (2018). \url{https://doi.org/10.1038/s41598-017-18635-w}.}

\leavevmode\vadjust pre{\hypertarget{ref-ritter2001}{}}%
\CSLLeftMargin{{[}5{]} }
\CSLRightInline{G. Ritter, L. Cohen, C. Williams, E. Richards, L. Old, S. Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}. {61} (2001) 6851--6859.}

\leavevmode\vadjust pre{\hypertarget{ref-roth2017}{}}%
\CSLLeftMargin{{[}6{]} }
\CSLRightInline{E.M. Roth, A.C. Goldberg, A.L. Catapano, A. Torri, G.D. Yancopoulos, N. Stahl, A. Brunet, G. Lecorps, H.M. Colhoun, Antidrug antibodies in patients treated with alirocumab, New England Journal of Medicine. 376 (2017) 1589--1590. \url{https://doi.org/10.1056/nejmc1616623}.}

\leavevmode\vadjust pre{\hypertarget{ref-jones2018}{}}%
\CSLLeftMargin{{[}7{]} }
\CSLRightInline{J.D. Jones, H.E. Ramser, A.E. Woessner, K.P. Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology. 1 (2018). \url{https://doi.org/10.1038/s42003-018-0206-4}.}

\leavevmode\vadjust pre{\hypertarget{ref-skala2010}{}}%
\CSLLeftMargin{{[}8{]} }
\CSLRightInline{M.C. Skala, A. Fontanella, L. Lan, J.A. Izatt, M.W. Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics. 15 (2010) 011112. \url{https://doi.org/10.1117/1.3285584}.}

\leavevmode\vadjust pre{\hypertarget{ref-greening2018}{}}%
\CSLLeftMargin{{[}9{]} }
\CSLRightInline{G.J. Greening, K.P. Miller, C.R. Spainhour, M.D. Cato, T.J. Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express. 9 (2018) 2871. \url{https://doi.org/10.1364/boe.9.002871}.}

\leavevmode\vadjust pre{\hypertarget{ref-sio2016}{}}%
\CSLLeftMargin{{[}10{]} }
\CSLRightInline{T.T. Sio, P.J. Atherton, B.J. Birckhead, D.J. Schwartz, J.A. Sloan, D.K. Seisler, J.A. Martenson, C.L. Loprinzi, P.C. Griffin, R.F. Morton, J.C. Anders, T.J. Stoffel, R.E. Haselow, R.B. Mowat, M.A.N. Wittich, J.D. Bearden, R.C. Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer. 24 (2016) 3847--3855. \url{https://doi.org/10.1007/s00520-016-3213-3}.}

\leavevmode\vadjust pre{\hypertarget{ref-kamstra2015}{}}%
\CSLLeftMargin{{[}11{]} }
\CSLRightInline{J.I. Kamstra, P.U. Dijkstra, M. van Leeuwen, J.L.N. Roodenburg, J.A. Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology. 51 (2015) 548--555. \url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.}

\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers2008}{}}%
\CSLLeftMargin{{[}12{]} }
\CSLRightInline{E.-J. Wagenmakers, M. Lee, T. Lodewyckx, G.J. Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer New York, 2008: pp. 181--207. \url{https://doi.org/10.1007/978-0-387-09612-4_9}.}

\leavevmode\vadjust pre{\hypertarget{ref-gueorguieva2004}{}}%
\CSLLeftMargin{{[}13{]} }
\CSLRightInline{R. Gueorguieva, J.H. Krystal, Move over {ANOVA}, Archives of General Psychiatry. 61 (2004) 310. \url{https://doi.org/10.1001/archpsyc.61.3.310}.}

\leavevmode\vadjust pre{\hypertarget{ref-schober2018}{}}%
\CSLLeftMargin{{[}14{]} }
\CSLRightInline{P. Schober, T.R. Vetter, Repeated measures designs and analysis of longitudinal data, Anesthesia {\&} Analgesia. 127 (2018) 569--575. \url{https://doi.org/10.1213/ane.0000000000003511}.}

\leavevmode\vadjust pre{\hypertarget{ref-pinheiro2006}{}}%
\CSLLeftMargin{{[}15{]} }
\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science; Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}

\leavevmode\vadjust pre{\hypertarget{ref-vishwanath2009}{}}%
\CSLLeftMargin{{[}16{]} }
\CSLRightInline{K. Vishwanath, H. Yuan, W.T. Barry, M.W. Dewhirst, N. Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia. 11 (2009) 889--900. \url{https://doi.org/10.1593/neo.09580}.}

\leavevmode\vadjust pre{\hypertarget{ref-dennis2019}{}}%
\CSLLeftMargin{{[}17{]} }
\CSLRightInline{B. Dennis, J.M. Ponciano, M.L. Taper, S.R. Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}, Frontiers in Ecology and Evolution. 7 (2019). \url{https://doi.org/10.3389/fevo.2019.00372}.}

\leavevmode\vadjust pre{\hypertarget{ref-wang2019}{}}%
\CSLLeftMargin{{[}18{]} }
\CSLRightInline{B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (2019) e100081. \url{https://doi.org/10.1136/gpsych-2019-100081}.}

\leavevmode\vadjust pre{\hypertarget{ref-liu2010}{}}%
\CSLLeftMargin{{[}19{]} }
\CSLRightInline{C. Liu, T.P. Cripe, M.-O. Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy. 18 (2010) 1724--1730. \url{https://doi.org/10.1038/mt.2010.127}.}
